Exception Handling Bug Hazards in Android

### 2015-10-12 SPRINGER_JOURNAL/msr_journal.tex
addressing Christpph and Arie's suggestions

\begin{abstract}
% Abstract somewhat based on Nature guidelines, 
% https://twitter.com/TrevorABranch/status/620699527486373888/photo/1?ref_src=twsrc%5Etfw
%
% Basic intro (1-2 sentences)
Adequate handling of exceptions has proven difficult for many software engineers.
% More detailed background (2-3 sentences)
Mobile app developers in particular, have to cope with 
compatibility, middleware, memory constraints, and battery restrictions.
% Clear sentence with general problem studied (1 sentence)
The goal of this paper is to obtain a thorough understanding of common exception handling
\emph{bug hazards} that app developers face.
% (method 1)
To that end, we first provide a detailed empirical study of over 6,000 
Java exception stack traces 
we extracted from over 600 open source Android projects.
% Main result 1 / new knowledge
Key insights from this study include common causes for system crashes, and
common chains of wrappings between checked and unchecked exceptions.
% (method 2)
Furthermore, we provide a survey with 70 developers involved
in at least one of the projects analyzed.
The results corroborate the stack trace findings, and indicate
that developers are unaware of frequently occurring undocumented exception handling behavior.
% General context and implications
Overall, the findings of our study call for tool support to help developers
understand their own and third party exception handling and wrapping logic.

@@

The exploratory survey was conducted with 70 Android developers involved in one or more of the 
GitHub Android projects whose issues were mined. This survey reveals that only few developers (3\% of respondents)
 knew about the undocumented checked exceptions signaled by native C code of the Android platform. 
Moreover, most of the developers recognized that cross-type exception wrappings may
negatively impact the application robustness. The uncaught exceptions due to errors in programming 
logic, e.g. the NullPointerException, were identified by most developers (68\%) as the first or second main cause 
of application crashes.

@@

 The NullPointerException was mainly signaled inside the application code (50\%) and the Android platform (31.5\%),
 although we could also find the NullPointerException being signaled by third-party libraries (16.3\%). 
Regarding reusable code (e.g., libraries and frameworks), there is no consensus whether it is a good or a bad practice to 
explicitly throw a NullPointerException. Some prefer to encapsulate such an exception in
an instance of IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
can be signaled by an API expecting a non-null argument.The high prevalence of NullPointerException is aligned with the
findings of other 
research~\cite{kim2013predicting,fraser20131600,csallner2004jcrasher,kechagia2014}. 
For instance, Kechagia and Spinellis showed that the NullPointerException was the  
most reported exception in the crash reports sent to BugSense (a bug report 
management service for Android applications)  \cite{kechagia2014}.
Other research on robustness testing~\cite{maji2012empirical,csallner2004jcrasher} shows that most of the automatically 
detected bugs were due to NullPointerException and exceptions
implicitly signaled by the Java 
environment due to programming mistakes or resource limitations
 (as the ones found in our study).

@@

\emph{\textbf{Identifying the Concerns Related to Root Exceptions.}} To get a broader view of the root exceptions of stack traces,
we performed a manual inspection in order to identify the underlying
concerns related to the most frequently reported root exceptions.
Besides the exceptions related to programming mistakes mentioned before, we also looked for exceptions related to concerns that are known as sources of faults in mobile development: concurrency~\cite{ama2012} backward compatibility~\cite{McDon13}, security~\cite{enck2011study,was2010} and resource management (IO, Memory, Battery)~\cite{Zhang12}. Since it is infeasible to inspect the code responsible for throwing every exception reported in this study,
the concern identification for each exception was based on intended
meaning of the particular exception type, as defined in 
its Javadoc documentation and in the Java specification. 
For example: (i) an instance of ArrayOutOfBoundException  
refers to a programming mistake according to its Javadoc; and (ii) the Java specification lists all 
exceptions related to backward compatibility~\cite{javaback}, such as
InstantiationError, VerifyError, and IllegalAccessError.

@@

%////// PREVENT NULL POINTER /////
Developers were also asked about what they usually do to prevent NullPointerExceptions from happening when their app is in production. Most of them mentioned including null-checks in the application code (59\%), mostly when checking the return of a method and as method guard conditions. Many respondents (39\%) also answered that instead of just sprinkling the code with null-checks when a NullPointerException happens they investigate the cause of the null reference and work to fix it, instead of just adding a null check. \emph{Determine why the object was null and attempt to fix this situation. In the case of external API calls which return null, then check for null (the quick-and-dirty way). For internal calls, use @NotNull and @Nullable annotations to provided more guidance on when an object ``may be'' and ``should never be'' null." [D42]}

@@

This question was motivated by the fact that during the mining study we could identify that many checked exceptions were just wrapped in a runtime exception and re-thrown, as a way to bypass any kind of handling. When asked about how to deal with a checked exception signaled by a method (invoked in the context of the onPause() activity method), most respondents (63\%) said that they should add a try-catch block surrounding the method invocation. One of the respondents mentioned that \emph{``If a method signature declares that it throws an exception, it means that the caller should handle the exception."[D29]}; \emph{``because it is checked exception, you should expect it to happen rather than just crash"[D70]}. However, most of the respondents also said that to deal with a method that is signaling a runtime exception we should surround such a method with a try-catch block (59\%).

@@

@@

\begin{table}
\scriptsize
\centering
\begin{tabular}{lrr}
\hline
\bfseries{Top Ways of Handling Checked Exceptions} & \bfseries{\#} & \bfseries{\%} \\
\hline
Add a try-catch block	& 25 &	63\% \\
Present error message	& 7 &	18\% \\
Involve the user in solving	& 5 &	13\% \\
Retry  &	5 &	13\% \\
Investigate the cause &	3 &	8\% \\
Same way as runtime	& 3 &	8\% \\
Add a throws declaration / pass upstream &	2 &	5\% \\
%Swallow	& 1 &	3\% \\
%Toast	 & 1 & 3\% \\
\hline
\end{tabular}
\caption{Top ways of dealing with a checked exception signaled in the context of an Activity method -- 40 non-empty responses. }
\label{tab:handlingruntime}
\end{table}

@@

Many respondents also said that they used testing approaches to prevent crashes -- few of them mentioned crowd testing (3\%), an emerging approach in testing apps. Moreover, null checks were also mentioned by 15\% of the developers as a way to prevent crashes. 
The adoption of coding styles was also pointed out by 14\% of respondents as a way to prevent crashes. Finally, crash report tools (e.g. ACRA~\cite{Acra14}) were reported as a way to prevent crashes. Once such tools notify developers about current crashes, the application can be fixed and hence future crashes (caused by what was fixed) could be prevented. Table~\ref{tab:topprevent} presents the themes for the strategies for crash prevention mentioned by developers.

@@

Another threat relates to the fact that parts of our analysis 
are based on the availability of stack traces in issues reported on GitHub and Google Code projects. 
In using these datasets, we make an underlying assumption: the stack traces reported in issues are 
representative of valid crash information of the applications. 
One way to mitigate this threat would be to access the full 
set of crash data per application. Although some services exist 
to collect crash data from mobile applications~\cite{BugSe14,BugSn14,Googl14,Acra14},
they do not provide open access to the crash reports of their client applications.
In our study, we mitigated this threat by manually inspecting
the source code associated with a subset of the reported exception stack traces.
This subset comprises the stack traces related to the main findings 
of the study (e.g., ``undocumented runtime and checked exceptions'',
and ``cross-type wrappings'').

@@

\textit{Analysis and Use of Stack Trace Information.} Several papers have
investigated the use of stack trace information to support: bug classification
and clustering~\cite{wang2013improving, kim2011crash, dhaliwal2011classifying},
fault prediction models~\cite{kim2013predicting}, automated
bug fixing tools~\cite{sinha2009fault} and also the analysis of Android APIs~\cite{kechagia2014}. 
Kim et al.~\cite{kim2011crash} use an
aggregated form of multiple stack traces available in crash reports to detect
duplicate crash reports and to predict if a given crash will be fixed. Dhaliwal
et al.~\cite{dhaliwal2011classifying} proposed a crash grouping approach that
can reduce bug fixing time by approximately 5\%. Wang et
al.~\cite{wang2013improving} propose an approach to identify correlated crash
types and describe a fault localization method to locate and rank files related
to the bug described in a stack trace. Schr\"oter et al.~\cite{schroter2010stack}
conducted an empirical study on the usefulness of stack traces for bug fixing
and showed that developers fixed the bugs faster when failing stack traces were
included in bug issues.  In a similar study, Bettenburg et
al.~\cite{bettenburg2008makes} identify stack traces as the second most stack
trace feature for developers.  Sinha et al.~\cite{sinha2009fault} proposed an
approach that uses stack traces to guide a dataflow analysis for locating and
repairing faults that are caused by the implicitly signaled exceptions. Kim
at al.~\cite{kim2013predicting} proposed an approach to predict the
crash-proneness of methods based information extracted from stack traces and
methods' bytecode operations.  They observed that most of the stack traces were
related to NullPointerException and other implicitly thrown exceptions had
the higher prevalence in the analyzed set of stacks. Kechagia and Spinellis~\cite{kechagia2014}
examined the stack traces embedded in crash reports sent by 1,800 Android apps 
to a crash report management service (i.e., BugSense). They found that 19\% of such stack traces
were caused by unchecked and undocumented exceptions thrown by methods defined in the 
Android API (level 15). Our work differs from Kechagia and Spinellis since it is based on
stack traces mined from issues reported by open source developers on GitHub and Google Code.  
Moreover, our study mapped the origin of each exception 
(i.e., libraries, the Android platform or the application itself) and investigated
the adoption of best practices based on the analysis of stack trace information.
Our work also identified the type of each exception mined from issues
(classifying them as Error, Runtime or Checked) based on the source code
analysis of the exception hierarchy and analyzed the exception wrappings that can 
happen during the exception propagation.  Such analysis revealed intriguing 
\emph{bug hazards} such as the cross-type exception wrappings not discussed in previous works.

>>>>

\begin{abstract}
% Abstract somewhat based on Nature guidelines, 
% https://twitter.com/TrevorABranch/status/620699527486373888/photo/1?ref_src=twsrc%5Etfw
%
% Basic intro (1-2 sentences)
Adequate handling of exceptions has proven difficult for many software engineers.
% More detailed background (2-3 sentences)
Mobile app developers in particular, have to cope with 
compatibility, middleware, memory constraints, and battery restrictions.
% Clear sentence with general problem studied (1 sentence)
The goal of this paper is to obtain a thorough understanding of common exception handling
\emph{bug hazards} that app developers face.
% (method 1)
To that end, we first provide a detailed empirical study of over 6,000 
Java exception stack traces 
we extracted from over 600 open source Android projects.
% Main result 1 / new knowledge
Key insights from this study include common causes for system crashes, and
common chains of wrappings between checked and unchecked exceptions.
% (method 2)
Furthermore, we provide a survey with 71 developers involved
in at least one of the projects analyzed.
The results corroborate the stack trace findings, and indicate
that developers are unaware of frequently occurring undocumented exception handling behavior.
% General context and implications
Overall, the findings of our study call for tool support to help developers
understand their own and third party exception handling and wrapping logic.

@@

The exploratory survey was conducted with 71 Android developers involved in one or more of the 
GitHub Android projects whose issues were mined. This survey reveals that only few developers (3\% of respondents)
 knew about the undocumented checked exceptions signaled by native C code of the Android platform. 
Moreover, most of the developers recognized that cross-type exception wrappings may
negatively impact the application robustness. The uncaught exceptions due to errors in programming 
logic, e.g. the NullPointerException, were identified by most developers (68\%) as the first or second main cause 
of application crashes.

@@

 The NullPointerException was mainly signaled inside the application code (50\%) and the Android platform (31.5\%),
 although we could also find the NullPointerException being signaled by third-party libraries (16.3\%). 
Regarding reusable code (e.g., libraries and frameworks), there is no consensus whether it is a good or a bad practice to 
explicitly throw a NullPointerException. Some prefer to encapsulate such an exception in
an instance of IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
can be signaled by an API expecting a non-null argument.The high prevalence of NullPointerException is aligned with the
findings of other 
research~\cite{kim2013predicting,fraser20131600,csallner2004jcrasher,kechagia2014}. 
For instance, Kechagia and Spinellis showed that the NullPointerException was the  
most reported exception in the crash reports sent to BugSense~\footnote{https://www.bugsense.com/} (a bug report 
management service for Android applications)~\cite{kechagia2014}.
Other research on robustness testing~\cite{maji2012empirical,csallner2004jcrasher} shows that most of the automatically 
detected bugs were due to NullPointerException and exceptions
implicitly signaled by the Java 
environment due to programming mistakes or resource limitations
 (as the ones found in our study).

@@

\emph{\textbf{Identifying the Concerns Related to Root Exceptions.}} To get a broader view of the root exceptions of stack traces,
we performed a manual inspection in order to identify the underlying
concerns related to the most frequently reported root exceptions.
Besides the exceptions related to programming mistakes mentioned before, we also looked for exceptions related to concerns that are known as sources of faults in mobile development: concurrency~\cite{ama2012} backward compatibility~\cite{McDon13}, security~\cite{enck2011study,was2010} and resource management (IO, Memory, Battery)~\cite{Zhang12}. Since it is infeasible to inspect the code responsible for throwing every exception reported in this study,
the concern identification for each exception was based on intended
meaning of the particular exception type, as defined in 
its Javadoc documentation and in the Java specification. 
For example: (i) an instance of ArrayOutOfBoundException  
refers to a programming mistake according to its Javadoc; and (ii) the Java specification lists all 
exceptions related to backward compatibility~\footnote{http://docs. oracle.com/javase/specs/jls/se7/html/jls-13.html}, such as
InstantiationError, VerifyError, and IllegalAccessError.

@@

%////// PREVENT NULL POINTER /////
Developers were also asked about what they usually do to prevent NullPointerExceptions from happening when their app is in production. Table~ref{tab:preventnull} presents the ways for preventing NullPointerException in Android apps most cited by developers. Most of the developers mentioned including null-checks in the application code (59\%), mostly when checking the return of a method and as method guard conditions. Many respondents (39\%) also answered that instead of just sprinkling the code with null-checks when a NullPointerException happens they investigate the cause of the null reference and work to fix it, instead of just adding a null check. \emph{Determine why the object was null and attempt to fix this situation. In the case of external API calls which return null, then check for null (the quick-and-dirty way). For internal calls, use @NotNull and @Nullable annotations to provided more guidance on when an object ``may be'' and ``should never be'' null." [D42]}

@@

This question was motivated by the fact that during the mining study we could identify that many checked exceptions were just wrapped in a runtime exception and re-thrown, as a way to bypass any kind of handling. When asked about how to deal with a checked exception signaled by a method (invoked in the context of the onPause() activity method), most respondents (63\%) said that they should add a try-catch block surrounding the method invocation - see Table~\ref{tab:handlingchecked}. One of the respondents mentioned that \emph{``If a method signature declares that it throws an exception, it means that the caller should handle the exception."[D29]}; \emph{``because it is checked exception, you should expect it to happen rather than just crash"[D70]}. However, most of the respondents also said that to deal with a method that is signaling a runtime exception we should surround such a method with a try-catch block (59\%) - see Table~\ref{tab:handlingruntime}.

@@

\begin{table}
\scriptsize
\centering
\begin{tabular}{lrr}
\hline
\bfseries{Top Ways of Handling Checked Exceptions} & \bfseries{\#} & \bfseries{\%} \\
\hline
Add a try-catch block	& 25 &	63\% \\
Present error message	& 7 &	18\% \\
Involve the user in solving	& 5 &	13\% \\
Retry  &	5 &	13\% \\
Investigate the cause &	3 &	8\% \\
Same way as runtime	& 3 &	8\% \\
Add a throws declaration / pass upstream &	2 &	5\% \\
%Swallow	& 1 &	3\% \\
%Toast	 & 1 & 3\% \\
\hline
\end{tabular}
\caption{Top ways of dealing with a checked exception signaled in the context of an Activity method -- 40 non-empty responses. }
\label{tab:handlingchecked}
\end{table}

@@

@@

Many respondents also said that they used testing approaches to prevent crashes -- few of them mentioned crowd testing (3\%), an emerging approach in testing apps. Moreover, null checks were also mentioned by 15\% of the developers as a way to prevent crashes. 
The adoption of coding styles was also pointed out by 14\% of respondents as a way to prevent crashes. Finally, crash report tools (e.g. ACRA~\footnote{code.google.com/p/acra/}) were reported as a way to prevent crashes. Once such tools notify developers about current crashes, the application can be fixed and hence future crashes (caused by what was fixed) could be prevented. Table~\ref{tab:topprevent} presents the themes for the strategies for crash prevention mentioned by developers.

%~\cite{Acra14}
\bigskip

@@

Another threat relates to the fact that parts of our analysis 
are based on the availability of stack traces in issues reported on GitHub and Google Code projects. 
In using these datasets, we make an underlying assumption: the stack traces reported in issues are 
representative of valid crash information of the applications. 
One way to mitigate this threat would be to access the full 
set of crash data per application. Although some services exist 
to collect crash data from mobile applications (e.g., ACRA~\footnote{code.google.com/p/acra/},  Google analytics~\footnote{https://www.google.com/analytics/}, Bugsense, Bugsnag~\footnote{https://bugsnag.com/}),
they do not provide open access to the crash reports of their client applications.
In our study, we mitigated this threat by manually inspecting
the source code associated with a subset of the reported exception stack traces.
This subset comprises the stack traces related to the main findings 
of the study (e.g., ``undocumented runtime and checked exceptions'',
and ``cross-type wrappings'').

@@

\textit{Analysis and Use of Stack Trace Information.} Several papers have
investigated the use of stack trace information to support: bug classification
and clustering~\cite{wang2013improving, kim2011crash, dhaliwal2011classifying},
fault prediction models~\cite{kim2013predicting}, automated
bug fixing tools~\cite{sinha2009fault} and also the analysis of Android APIs~\cite{kechagia2014}. 
Kim et al.~\cite{kim2011crash} use an
aggregated form of multiple stack traces available in crash reports to detect
duplicate crash reports and to predict if a given crash will be fixed. Dhaliwal
et al.~\cite{dhaliwal2011classifying} proposed a crash grouping approach that
can reduce bug fixing time by approximately 5\%. Wang et
al.~\cite{wang2013improving} propose an approach to identify correlated crash
types and describe a fault localization method to locate and rank files related
to the bug described in a stack trace. Schr\"oter et al.~\cite{schroter2010stack}
conducted an empirical study on the usefulness of stack traces for bug fixing
and showed that developers fixed the bugs faster when failing stack traces were
included in bug issues.  In a similar study, Bettenburg et
al.~\cite{bettenburg2008makes} identify stack traces as the second most relevant
 feature of good bug reports.  Sinha et al.~\cite{sinha2009fault} proposed an
approach that uses stack traces to guide a dataflow analysis for locating and
repairing faults that are caused by the implicitly signaled exceptions. Kim
at al.~\cite{kim2013predicting} proposed an approach to predict the
crash-proneness of methods based information extracted from stack traces and
methods' bytecode operations.  They observed that most of the stack traces were
related to NullPointerException and other implicitly thrown exceptions had
the higher prevalence in the analyzed set of stacks. Kechagia and Spinellis~\cite{kechagia2014}
examined the stack traces embedded in crash reports sent by 1,800 Android apps 
to a crash report management service (i.e., BugSense). They found that 19\% of such stack traces
were caused by unchecked and undocumented exceptions thrown by methods defined in the 
Android API (level 15). Our work differs from Kechagia and Spinellis since it is based on
stack traces mined from issues reported by open source developers on GitHub and Google Code.  
Moreover, our study mapped the origin of each exception 
(i.e., libraries, the Android platform or the application itself) and investigated
the adoption of best practices based on the analysis of stack trace information.
Our work also identified the type of each exception mined from issues
(classifying them as Error, Runtime or Checked) based on the source code
analysis of the exception hierarchy and analyzed the exception wrappings that can 
happen during the exception propagation.  Such analysis revealed intriguing 
\emph{bug hazards} such as the cross-type exception wrappings not discussed in previous works.
